---
title: "Multivariate Data Analysis Project"
output:
  html_document: default
  pdf_document: default
date: "2022-12-14"
---
```{r include=FALSE}
#Packages
#install.packages("skimr")
library("skimr")
#install.packages("car")
library("car")
#install.packages("ggplot2")
library(ggplot2)
#install.packages("factoextra")
library(factoextra)
#install.packages("AMR")
library(AMR)
#install.packages("devtools")
library(devtools)
#install.packages("corrplot")
library(corrplot)
#install.packages("dplyr")
library(dplyr)
#install.packages("viridis")
library(viridis)
#install.packages("hrbrthemes")
library(hrbrthemes)
#install.packages("tidyverse")
library(tidyverse) #data manipulation
#install.packages("cluster")
library(cluster) #clustering algorithms
#install.packages("dendextend")
library(dendextend) #for comparing 2 dendograms
#install.packages("laGP")
library(laGP) #squared Euclidean distance
#install.packages("gridExtra")
library(gridExtra)

#install_github("vqv/ggbiplot")
require(ggbiplot)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.Preliminary analysis of the data 
```{r}
data_7 <- read.csv("data_7.csv", header = TRUE)
head(data_7)
```
```{r}
# Dimension
dim(data_7)
```
## Data summary
```{r}
skim(data_7)
```
## Boxplot
```{r message=FALSE, warning=FALSE}
#PreProcessing Data for BoxPloT
df_boxplot <- scale(data_7)
df_boxplot <- data.frame(df_boxplot)

df_boxplot$index <- 1:nrow(df_boxplot)

id_df_long <- reshape(data=df_boxplot, idvar="index",
                         varying = c("HGLB","VGLOB", "GBR", "LINF", "NEUTR", "CCSER"),
                         v.name=c("value"),
                         times=c("HGLB","VGLOB", "GBR", "LINF", "NEUTR", "CCSER"),
                         new.row.names = 1:1000,
                         direction="long")

names(id_df_long)[1] = "Individual"
names(id_df_long)[2] = "Variable"

#BoxPlot
Boxplot <- 
  ggplot(id_df_long, aes(x = factor(Variable), y = value, fill = Variable)) + 
  geom_boxplot(notch = TRUE) +
  labs(
    x = "Variable",
    y = "Standardized Value") +
  stat_summary(
    aes(label = round(stat(y), 1)),
    geom = "text", 
    fun.y = function(y) { o <- boxplot.stats(y)$out; if(length(o) == 0) NA else o },
    hjust = -1) 



#Which Individuals are ouliers?
findoutlier <- function(x) {return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))}

df <- id_df_long %>%
        group_by(Variable) %>%
        mutate(outlier = ifelse(findoutlier(value), Individual, NA))


Boxplot
```
*Outliers:*
HGLB: 12, 23\
VGLOB: 1, 12, 23, 38 \
GBR: 23\
LINF: 31 45 47\
NEUTR: - \
CCSER: 10\

# 2. Conducting principal component analysis
## Determine the correlation matrix
```{r}
cor_matrix <- cor(data_7)
corrplot(cor_matrix, tl.col = "black", tl.srt = 45, bg = "white",
         type = "upper", method = "number")
```
## Obtain eigenvalues and eigenvectors
```{r}
eigen_data_7 <- eigen(cor_matrix)
eigen_data_7
```
## Performing PCA
```{r}
pca_data_7 <- princomp(data_7, cor = TRUE)
print(summary(pca_data_7), loadings = TRUE)
```
## ScreePlot
```{r}
scree_plot <- fviz_screeplot(pca_data_7, addlabels = TRUE)
```
## Component Matix (loadings)
```{r}
component_matrix <- cor(data_7, pca_data_7$scores)
component_matrix
```
## Proportion of (standardized) population standard deviation (sd)  for  each PC
```{r}
# Proportion of (standardized) population standard deviation (sd) due to the 1st pc:
sqrt(eigen_data_7$values[1]/6)
# Proportion of (standardized) population standard deviation (sd) due to the 2nd pc:
sqrt(eigen_data_7$values[2]/6)
# Proportion of (standardized) population standard deviation (sd) due to the 3rd pc:
sqrt(eigen_data_7$values[3]/6)
```
## Contribution of each important variable for each principal component considered:
```{r}
# First PC
# Variables  VGLOB, GBR, LINF
all <- c( (component_matrix[2,1]/sqrt(eigen_data_7$values[1]))^2, (component_matrix[3,1]/sqrt(eigen_data_7$values[1]))^2, (component_matrix[4,1]/sqrt(eigen_data_7$values[1]))^2)
all
# Second PC
# Variables HGLB and NEUTR
all <- c((component_matrix[1,2]/sqrt(eigen_data_7$values[2]))^2, (component_matrix[5,2]/sqrt(eigen_data_7$values[2]))^2)
all
# Third PC
# Variable CCSER
all <- (component_matrix[6,3]/sqrt(eigen_data_7$values[3]))^2
all
```
## Graphical representation of principal components with Biplot:
```{r}
index <- 1:nrow(data_7) # To insert labels in Biplots
index

help("ggbiplot")
bplot_pc1_pc2= ggbiplot(pca_data_7, choices = c(1,2), labels = index)
bplot_pc1_pc3= ggbiplot(pca_data_7, choices = c(1,3), labels = index)
bplot_pc2_pc3= ggbiplot(pca_data_7, choices = c(2,3), labels = index)
```
```{r}
bplot_pc1_pc2
```
```{r}
bplot_pc1_pc3
```
```{r}
bplot_pc2_pc3
```
## Scores 
*pc1*
$p.c.1 = -0.424HGLB-0.446VGLOB-0.263GBR-0.454LINF-0.303NEUTR-0.073CCSER$
*pc2*
$p.c.2 = 0.561HGLB+0.528VGLOB-0.387GBR-0.267LINF-0.425NEUTR-0.069CCSER$
*pc3*
$p.c.3 = 0.149HGLB-0.087VGLOB+0.051GBR-0.165LINF+0.295NEUTR-0.923CCSER$

# 3. Cluster analysis
```{r}
data_7_scaled <- scale(data_7)
```
## Euclidean Distance (Scaled Data)
```{r}
#Dissimilarity matrix: euclidean distance
dist_data_7_scaled = dist(data_7_scaled, method="euclidean")
```
## Single Linkage Method
```{r}
hc_single_scaled = agnes(dist_data_7_scaled, method = 'single')
hc_single_scaled$merge
```
### Order
```{r}
hc_single_scaled$order
```
### Agglomerative Coefficient
```{r}
hc_single_scaled$ac
```
### Dendrogram (k=4)
```{r}
dend_single_scaled = as.dendrogram(hc_single_scaled)
plot(dend_single_scaled, main = "Single linkage")
rect.hclust(hc_single_scaled, k = 4, border = 2:4)
```
## Complete Linkage Method
```{r}
hc_complete_scaled = agnes(data_7_scaled, method = 'complete')
hc_complete_scaled$merge
```
### Agglomerative Coefficient
```{r}
hc_complete_scaled$ac
```
### Dendrogram (k=4; k=10)
```{r}
dend_complete_scaled = as.dendrogram(hc_complete_scaled)
par(mfrow=c(2,1))
plot(dend_complete_scaled, main = "Complete linkage - 4 clusters")
rect.hclust(hc_complete_scaled, k = 4, border = 2:4)
plot(dend_complete_scaled,main = "Complete linkage - 10 clusters")
rect.hclust(hc_complete_scaled, k = 10, border = 2:4)
```
## Average linkage Method
```{r}
hc_average_scaled = agnes(data_7_scaled, method = 'average')
hc_average_scaled$merge
```
### Agglomerative Coefficient
```{r}
hc_average_scaled$ac
```
### Dendrogram (k=4; k=10)
```{r}
dend_average_scaled = as.dendrogram(hc_average_scaled)
par(mfrow=c(2,1))
plot(dend_average_scaled, main = "Average linkage - 4 clusters")
rect.hclust(hc_average_scaled, k = 4, border = 2:4)
plot(dend_average_scaled, main = "Average linkage - 10 clusters")
rect.hclust(hc_average_scaled, k = 10, border = 2:4)
```
## Ward Method
```{r}
hc_ward_scaled = agnes(data_7_scaled, method = 'ward')
hc_ward_scaled$merge
```
### Agglomerative Coefficient
```{r}
hc_ward_scaled$ac
```
### Dendrogram (k=4; k=10)
```{r}
dend_ward_scaled = as.dendrogram(hc_ward_scaled)
par(mfrow=c(2,1))
plot(dend_ward_scaled, main = "Ward method - 4 clusters")
rect.hclust(hc_ward_scaled, k = 4, border = 2:4)
plot(dend_ward_scaled, main = "Ward method - 10 clusters")
rect.hclust(hc_ward_scaled, k = 10, border = 2:4)
```
## Dendrograms for all methods
```{r}
par(mfrow=c(2,2))
plot(dend_single_scaled,main = "Single linkage")
rect.hclust(hc_single_scaled, k = 4, border = 2:4)
plot(dend_complete_scaled,main = "Complete linkage")
rect.hclust(hc_complete_scaled, k = 4, border = 2:4)
plot(dend_average_scaled,main = "Average linkage")
rect.hclust(hc_average_scaled, k = 4, border = 2:4)
plot(dend_ward_scaled,main = "Ward method")
rect.hclust(hc_ward_scaled, k = 4, border = 2:4)
```
## CLusters Interpretation
### Plots 
```{r}
# Individuals and clusters
vect <- cbind(cutree(dend_ward_scaled, 4))

colnames(vect)[1] <- "Cluster"

Clusters <- data.frame(vect)

id_cluster_df <- cbind(Clusters, data_7_scaled)

# Wide to Long for plots
id_cluster_df_long <- reshape(data=id_cluster_df, idvar="Cluster",
                         varying = c("HGLB","VGLOB", "GBR", "LINF", "NEUTR", "CCSER"),
                         v.name=c("value"),
                         times=c("HGLB","VGLOB", "GBR", "LINF", "NEUTR", "CCSER"),
                         new.row.names = 1:1000,
                         direction="long")

# Create Plots
## Plot1
clusters_plot <- ggplot(data = id_cluster_df_long, aes(x = id_cluster_df_long$time, y = id_cluster_df_long$value)) +
  geom_bar(stat = "identity") +
  facet_grid(. ~id_cluster_df_long$Cluster)  +
  ylim(-5,5) +
  theme(axis.text.x = element_text(angle = 90)) +
  #geom_text(aes(label = id_cluster_df_long$value), fontface = "bold", vjust = 1.5, colour = "white", size = 4) +
  labs(x = "\n hematological variables", y = "standardized value\n", title = "\n hematological variables by cluster\n") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="royalblue4", size = 12),
        axis.title.y = element_text(face="bold", colour="royalblue4", size = 12),
        legend.title = element_text(face="bold", size = 10),
        strip.background = element_rect(fill="ghostwhite", colour="royalblue4", size=.7),
        strip.text = element_text(face="bold", size=rel(.7))) 

## Plot 2
cluster_plot2 <- ggplot(data = id_cluster_df_long, aes(x = id_cluster_df_long$time, y = id_cluster_df_long$value))+ 
    geom_bar(position="dodge", stat="identity") +
    scale_fill_viridis(discrete = T, option = "E") +
    ggtitle("hematological variables by cluster") +
    facet_wrap(~id_cluster_df_long$Cluster) +
    theme_ipsum() +
    theme(legend.position="none")+
    xlab("")+
    ylab("")+
    ylim(0,5)
```
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Call Plots
clusters_plot
```

## Data Frames with individuals
```{r}
#data frames for each cluster
vect <- cbind(cutree(dend_ward_scaled,4))

colnames(vect)[1] <- "Cluster"

Clusters <- data.frame(vect)

id_cluster_df <- cbind(Clusters,data_7)

cluster1 <- subset(id_cluster_df, subset = id_cluster_df$Cluster==1)
cluster2 <- subset(id_cluster_df, subset = id_cluster_df$Cluster==2)
cluster3 <- subset(id_cluster_df, subset = id_cluster_df$Cluster==3)
cluster4 <- subset(id_cluster_df, subset = id_cluster_df$Cluster==4)
```
```{r}
cluster1
```
```{r}
cluster2
```
```{r}
cluster3
```
```{r}
cluster4
```


```{r}
plot(dend_ward_scaled, main = "Ward method - 10 clusters")
rect.hclust(hc_ward_scaled, k = 10, border = 2:4)
```









